---
title: "Building and Distributing R Packages using GitHub"
subtitle: "Using GitHub for Open-Source Analytics, Reporting, and Dissemination of Research"
author: "Damian Betebenner"
institute: >-
  Center for Assessment
  <a href="https://github.com/CenterForAssessment/NCME_2022_Demonstration/blob/main/Presentations/Building_and_Distributing_R_Packages_Using_GitHub.Rmd" class="github-corner" aria-label="View source on Github"> <svg width="80" height="80" viewBox="0 0 250 250" style="position: absolute; top: 0px; right: 0px; border: 0px;" aria-hidden="true"> <path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z" fill="#000000"></path> <path class="octo-arm" d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="#6e5494" style="transform-origin: 130px 106px;"></path> <path class="octo-body" d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="#6e5494"></path> </svg> </a>

date: "April 21st, 2022 (updated: `r toOrdinal::toOrdinalDate(Sys.Date())`)<br />NCME 2022 Annual Meeting"
output:
  cfaDocs::cfaXaringan:
    cfa_theme: 'cfa-b'
    include: !expr list(
      xaringanExtra = 'default',
      xExtra_use = c('share_again', 'tile_view', 'clipboard'))

---

layout: true

<!--  Add Footer Manually - might only be able to have ONE 'layout: true' slide  -->
<!--  this adds the footer to all slides, depends on footer class in css  -->

<div class="footer" style="left: 5%;"><img src="https://mirrors.creativecommons.org/presskit/icons/cc.svg" style = "max-width: 40%; opacity: 0.5;"></img></div>
<div class="footer" style="left: 7.5%;"><img src="https://mirrors.creativecommons.org/presskit/icons/by.svg" style = "max-width: 40%; opacity: 0.5;"></img></div>
<div class="footer" style="left: 40%;"><a class="footer" href="https://github.com/CenterForAssessment/packageSkeletonForR">R Packages with GitHub</a></div>


```{r metadata, echo=FALSE}
# creating HTML metadata to accompany the slides
library(metathis)
meta() %>%
  meta_general(
    description = "The first presentation of the NCME 2022 Demonstration Session on Using GitHub for Open-Source Analytics, Reporting, and Dissemination of Research",
    subject = "Building and Distributing R Packages Using GitHub",
    generator = "xaringan and remark.js",
    theme_color = "#b6e202"
  ) %>% 
  meta_name("github-repo" = "centerforassessment/NCME_2022_Demonstration") %>% 
  meta_social(
    title = "Building & Distributing R Packages Using GitHub • NCME 2022 • Damian Betebenner",
    url = "https://centerforassessment.github.io/NCME_2022_Demonstration/articles/presentations/Demonstration_Session_Presentations.html#using-github-to-build-distribute-r-packages",
    image = "https://centerforassessment.github.io/NCME_2022_Demonstration/presentations/Building_and_Distributing_R_Packages_Using_GitHub_files/TITLE_SLIDE_image.png",
    image_alt = "Title slide for Building & Distributing R Packages Using GitHub presentation",
    og_type = "article",
    og_locale = "en_US",
    og_author = "Damian Betebenner",
    twitter_card_type = "summary_large_image",
    twitter_creator = "@dbetebenner",
    twitter_site = "@NCIEA1"
  )
```

---
class: inverse, center, middle

# R
---
class: highlight-last-item

``` {css, echo=FALSE}
h1 > img {
  width: 10%
}
```

# ![](https://www.r-project.org/logo/Rlogo.svg)

--
- Open soure software environment originally developed for statistical analysis (but now useful for lots more)

--
- Has become one of the most prominent environmetns for performing and developing data analytics/statistics

--
- Applicable to many other things (this presentation and the website on which it is based was created using R)

--
- One of its greatest strengths is its extensibility

--
- Allows users to create and distribute customized R software

--
- Currently 19,002 package on CRAN and many more available only on GitHub (which is a big part of what I'm going to talk about)

--
- <hr />

---
class: inverse, center, middle

# GitHub

---
class: highlight-last-item

# GitHub

--

- GitHub is a code hosting platform for version control and collaboration

--

- Collaboration without coordination

--

- Often used for software development but is incredibly useful for data analytics

--

- <hr/>

---

class: inverse, center, middle

# GitHub

# +

# R

# =

# Superpower

---
class: highlight-last-item

# GitHub + R

--

- The combination of GitHub and R has been the foundation of my work for the last 15 years.

--

- The combination supports and improves:

--
    - Reproducibility of analytics
--
    - Collaboration amongst team members
--
    - Dissemination of analytics and findings
--

- If data analysis and reporting results is your career, I can't recommend the combination of GitHub + R enough.

--
- <hr/>

---
class: highlight-last-item

# R Packages

--

- R provides a simple to learn way of extending base R by creating packages

--

- The packages can be easily shared with other to share analytic or other extended R functionality

--

-   

---
class: highlight-last-item

# R Packages

--
- Packages can be shared and installed by others in a multide of ways:

--
  - Email a zipped bundle to another person who can install it. (very oldschool and clunky)

--
  - Submit it to CRAN and use `install.packages()` to install the package. (Good for "official versions/archiving")

--
  - Build it on GitHub and use `remotes::install_github()` or `devtools::install_github` to install off of GitHub. (Get version control in addition to package hosting)

--
- The definitive guide on R package construction is [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html) available on CRAN.

--
- <hr/>

---
class: highlight-last-item

# packageSkeletonForR

--

- We create a GitHub repo that is an R package skeleton for R

--

- [packageSkeletonForR](https://github.com/CenterForAssessment/packageSkeletonForR)

--

- The skeleton contains many of the pieces that you'll need to have in order to have a package

--

- The skeleton also contains several additional pieces (a website) describing the contents of the repo and how it works

--
- <hr/>

---

# packageSkeletonForR

```{r echo=FALSE, out.width="80%", fig.align="center"}
magick::image_read_pdf(path="Figs/packageSkeletonForR.pdf")
```

---
class: highlight-last-item

# R Package Contents

--
- A folder `R` that contains your R scripts/functions

--
- A folder `inst` that contains `CITATION` and `NEWS` files

--
- A folder `man` that contains the manual pages associated with your exported functions

--
- A folder `tests` that contains tests associated with your functions (optional but recommended)

--
- A folder `vignettes` that contains RMarkdown vignettes associated with your package (optional but recommended)

--
- A file `DESCRIPTION` that contains text describing the package

--
- A file `NAMESPACE` that specific what gets imported and exported to/from the package.

--
- <hr/>

---
class: highlight-last-item

# R Package Contents

--
- A Folder `.github` that contains a file allowing one to use GitHub to compile the R package and test for errors.

--
- A folder `docs` that contains the source code associated with the GitHub hosted website for the package (created by R package `packagePages` using the function `projectPages`)

--
- A file `.Rbuildignore` that specifies which files/folders to ignore when the package gets built.

--
- A file `.appveyor` that specific another continuous integration application (appveyor). Probably redundant with GitHub actions

--
- Files `README.md`, `NEWS.md` and `LICENSE.md` that are markdown based and provide information about the GitHub repo.

--
- A file `_pkgdown` which provides the structure for the dropdown menu of the website associated with the repo.

--
- <hr/>

---
class: highlight-last-item

# Package Installation

--

### To install the package

--

- `remotes::install_github("centerforassessment/packageSkeletonForR")`

--

- Changes to the package can be made via commits to the GitHub repo which become `pull requests`

--

- Multiple people can make commits with different permissions as <br/> to who can merge in `pull requests`

--
- <hr/>
